---
title: "riparian"
author: "Erik Rose"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{riparian}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Welcome to the `vignette` for the package `riparian`, a suite of tools designed to help Benton County Community Development staff analyze riparian cover extent and change in cover over time.  This document models the workflow of producing an annual monitoring report for the SRP program, and demonstrates how functions in the `riparian` package can make monitoring faster and more efficient for the researcher.

## Step One:  Site Plots

Benton County conducts riparian monitoring whenever new high resolution (1-foot) orthographic imagery becomes available.  Likely, your bosses have hired you because new imagery has just arrived.  In my year, the County collected collected imagery in S:/maps/OrthoPhotos, which each year of imagery in a separate folder.  If you are planning on doing a lot of operations on large raster data, then I suggest copying the folder containing the imagery onto an external drive.  This minimizes network traffic for others and speeds performance at your workstation.

The County currently monitors 54 randomly selected sites along priority corridors, and may add additional sites based upon future permit approvals.  The reason we look at the same 54 sites every year is because we are trying to detect change in cover.  If we looked at a different slew of sites each year, it would be hard to know whether an increase or decrease in cover were due to variation between the sites we selected, or whether vegetation removal or accrual occurred.  Using this method, we can circle individual trees that have gone missing from one year to the next.  Because the sites are randomly selected and representative of conditions in the priority corridor generally, we can be reasonably confident that changes in cover within the sample reflect conditions in the priority corridor.

Acces the polygon shapefile containing the 54 sampling boxes for each site by calling the command:

```{r}
data(samples, package = 'riparian')

# use the raster package to plot spatial data
raster::plot(samples[1,], main = 'Example Sampling Box')
```

The function for printing plots of sample sites using new orthographic imagery is called `plot_samples()`, and it takes three main arguments.  The first argument is called `in_path` and it is the path to the directory of orthographic imagery.  R uses the convention of forward slashes in path names (for instance the path to my working folder for this project is 'E:/Riparian' and the path to the 2018 imagery is 'E:/ortho2018').  The function expects a character string, so the path name needs to have single or double quotes around it.  The second argument is `out_path`, and specifies where the function will print the sample plots into.  The third argument, `samples` is the polygon shapefile of sampling boxes.  By default the function looks for an object called `samples` in the current environment, so if you have loaded the data `samples`, then you can omit the third argument.

```{r, eval=FALSE}
library(riparian)

# not run
# change the path names to use
plot_samples(in_path = 'E:/ortho2018',
             out_path ='E:/Riparian/sites2018')
```
```{r, echo=FALSE}
# example output
pic_path <- system.file('extdata/images', 'eg1.png', pacakge = 'riparian')
knitr::include_graphics(pic_path)
```

The function `plot_samples()` has additional optional arguments.  By default it will print an RGB plot and an NDVI plot for each site.  You can turn off one or the other by setting `print_rgb` or `print_ndvi` to FALSE, depicted below:

```{r, eval=FALSE}
# not run

# prints rgb only
plot_samples('E:/ortho2018', 'E:/Riparian/sites2018', print_ndvi = FALSE)

# prints ndvi only
plot_samples('E:/ortho2018', 'E:/Riparian/sites2018', print_rgb = F)
```


## Scoring Samples

Once you have printed sampling plots for all sites, it is time to score cover extent.

In 2018 I created an excel file called `samples2018.xlsx` to record cover extent for the 2018 annual monitoring report and the 5-year trends and status report.  In order for the functions in `riparian` to work correctly, it is important for you to format your record of observations exactly like the spreadsheet from 2018.  First, let's take a look at that spreadsheet by loading it into the workspace.  The file is located in the extdata folder in csv format.

```{r, eval=F}
library(data.table)

# construct the file path to csv
csv_path <- system.file('extdata', 'samples2018.csv', package = 'riparian')

# read the csv in as a data.table
obs <- fread(csv_path)

# examine results
str(obs)
```

You can also examine the data as a native R object, using the `data` function.

```{r}
obs <- data(samples2018, package = 'riparian')

str(obs)
```




