---
title: "riparian"
author: "Erik Rose"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{riparian}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Welcome to the `vignette` for the package `riparian`, a suite of tools designed to help Benton County Community Development staff analyze riparian cover extent and change in cover over time.  This document models the workflow of producing an annual monitoring report for the SRP program, and demonstrates how functions in the `riparian` package can make monitoring faster and more efficient for the researcher.

## Step One:  Site Plots

Benton County conducts riparian monitoring whenever new high resolution (1-foot) orthographic imagery becomes available.  Likely, your bosses have hired you because new imagery has just arrived.  In my year, the County collected collected imagery in S:/maps/OrthoPhotos, which each year of imagery in a separate folder.  If you are planning on doing a lot of operations on large raster data, then I suggest copying the folder containing the imagery onto an external drive.  This minimizes network traffic for others and speeds performance at your workstation.

The County currently monitors 54 randomly selected sites along priority corridors, and may add additional sites based upon future permit approvals.  The reason we look at the same 54 sites every year is because we are trying to detect change in cover.  If we looked at a different slew of sites each year, it would be hard to know whether an increase or decrease in cover were due to variation between the sites we selected, or whether vegetation removal or accrual occurred.  Using this method, we can circle individual trees that have gone missing from one year to the next.  Because the sites are randomly selected and representative of conditions in the priority corridor generally, we can be reasonably confident that changes in cover within the sample reflect conditions in the priority corridor.

Acces the polygon shapefile containing the 54 sampling boxes for each site by calling the command:

```{r}
data(samples, package = 'riparian')

# use the raster package to plot spatial data
raster::plot(samples[1,], main = 'Example Sampling Box')
```

The function for printing plots of sample sites using new orthographic imagery is called `plot_samples()`, and it takes three main arguments.  The first argument is called `in_path` and it is the path to the directory of orthographic imagery.  R uses the convention of forward slashes in path names (for instance the path to my working folder for this project is 'E:/Riparian' and the path to the 2018 imagery is 'E:/ortho2018').  The function expects a character string, so the path name needs to have single or double quotes around it.  The second argument is `out_path`, and specifies where the function will print the sample plots into.  The third argument, `samples` is the polygon shapefile of sampling boxes.  By default the function looks for an object called `samples` in the current environment, so if you have loaded the data `samples`, then you can omit the third argument.

```{r, eval=FALSE}
library(riparian)

# not run
# change the path names to use
plot_samples(in_path = 'E:/ortho2018',
             out_path ='E:/Riparian/sites2018')
```
```{r, echo=FALSE}
# example output
pic_path <- system.file('extdata/images', 'eg1.png', pacakge = 'riparian')
knitr::include_graphics(pic_path)
```

The function `plot_samples()` has additional optional arguments.  By default it will print an RGB plot and an NDVI plot for each site.  You can turn off one or the other by setting `print_rgb` or `print_ndvi` to FALSE, depicted below.  

If you have a new polygon object, perhaps as the result of generating new sampling sites, then set the argument `samples` to the polygon object.


```{r, eval=FALSE}
# not run

# prints rgb only
plot_samples('E:/ortho2018', 'E:/Riparian/sites2018', print_ndvi = FALSE)

# prints ndvi only
plot_samples('E:/ortho2018', 'E:/Riparian/sites2018', print_rgb = F)

library(rgdal)
setwd('path_to_shapefile')
polys <- readOGR('mysamples.shp')

plot_samples('E:/ortho2018', 'E:/Riparian/sites2018', samples = polys)
```




## Scoring Samples

Once you have printed sampling plots for all sites, it is time to score cover extent.

In 2018 I created an excel file called `samples2018.xlsx` to record cover extent for the 2018 annual monitoring report and the 5-year trends and status report.  In order for the functions in `riparian` to work correctly, it is important for you to format your record of observations exactly like the spreadsheet from 2018.  First, let's take a look at that spreadsheet by loading it into the workspace.  The file is located in the extdata folder in csv format.

```{r, eval=F}
library(data.table)

# construct the file path to csv
csv_path <- system.file('extdata', 'samples2018.csv', package = 'riparian')

# read the csv in as a data.table
obs <- fread(csv_path)

# examine results
str(obs)
```

You can also examine the data as a native R object, using the `data` function.

```{r}
data(samples2018, package = 'riparian')

head(samples2018, 10)
```

The spreadsheet has 53 columns, named `id`, `year`, `type` and then numbered 1:50.  Your spreadsheet must adhere to this format, using 53 columns named just so.  The `id` column refers to an individual site id assoiated with each site.  The `year` column specifies the year of observation.  The `type` column is for grouping sites into subsets, such as sites with active permits to track compared to random samples, and may be left blank without consequence.  The columns numbered 1:50 correspond to the numbers on the sample plots printed by `plot_samples()`.  When you are scoring observations, enter the cover score associated with the number on the sampling plot under the column with the same number in the spreadsheet.

The easiest way to adhere to the format requirement is to simply add new observations to the bottom of the 2018 spreadsheet, and save it under a new name.  You can convert the R object into a csv file using the command `write.csv(samples2018, 'myfile.csv')` and fill in your observations using Excel.

When you are done editing your file in Excel, export it into csv format, which is a tried and true way to share data between programs.  Read the csv into RStudio using the `data.table` package.

```{r, eval = F}
# not run
library(data.table)
setwd('path_to_file')

obs <- fread('myobs.csv')
```


### Subsetting Sites

The `riparian` package takes advantage of several features of the `data.table` package.  The `data.table` is a special class of `data.frame`, with unique syntax for subsetting data.

The first subsetting you will likely want to do is subset by year.  Do this by selecting the column `year` and setting it equal to the target value.  Place this logical test in brackets following the data.table name, and R will return only those observations that pass the test.  Note that logical tests use two equal signs (eg. `year == 2018`).

```{r}
library(data.table)
# print summary table of obs by year
samples2018[ , .N, by = year]

# subset year of interst
sub18 <- samples2018[year == 2018]

# very subset is correct
sub18[ , .N, by = year]
```

The `type` variable is a catch-all for you define and compare groups within the sample.  The 2018 dataset includes a type value `permit`, which refers to sites with active permits we selected for random monitoring.  We can split the dataset into two groups by using `==` and `!=` (not equal) to divide the data by `type`:

```{r}
# print table of obs by type
samples2018[ , .N, by = type]

# subset permit and nopermit sites
permit_sites <- samples2018[type == 'permit']
nopermit_sites <- samples2018[type != 'permit']

# examine results
permit_sites[ , .N, by = type]
nopermit_sites[ , .N, by = type]
```









